MS_Intro_Short

### 2023-10-16 Markdown/Manuscript/Child_RMDs/MS_Discussion.Rmd
missing tables

<!-- Talk about DARTH (https://doi.org/10.1007/s40273-019-00837-x )/ BCEA -->
Ethical practice is a core expectation of health researchers and computational methods underpin most quantitative research, yet an understanding of what constitutes ethical computational modelling practice in health is underdeveloped [@10.3389/fpubh.2017.00068]. We have addressed this gap with our proposed modeller and funder responsibilities, enabling model attributes and model implementation assessment criteria.

@@

As illustrated by Table \@ref(tab:checktb), we have developed a prototype software framework that can be used to implement youth mental health CHEMs that largely satisfy our TRU criteria. However, in its current form our software framework is too fragile to be anything more than a prototype for supporting the development needs other modelling teams and projects. We developed our prototype software framework was developed with the needs of only one group of developers in mind – ourselves. We currently lack the resources required to adequately implement strategies to target factors such as user enjoyment, usability, active user-community and supporting resources that influence adoption of software frameworks [@myllarniemi2018development].

Our prototype has a number of features that subsequent work to develop CHEM software frameworks may find useful to incorporate. Firstly, developing a software framework to work within an existing and widely used open source programming language such as R or python, can keep framework scope relatively narrow. This makes it more tractable to develop, maintain and learn, and can integrate other modelling tools written in that language (e.g., the dependency libraries we list in Table \@ref(tab:cpkgs)). Secondly, implementation that combines both object oriented and functional programming paradigms can avail of the modular and syntactical simplicity benefits of the former, while limiting needless bundling of code artefacts. Thirdly, a sensible trade-off needs to be found between transparent code implementation (which requires clear and sufficiently detailed documentation) and Agile Software Development (for which a foundational principle is prioritising the development of working code over writing documentation [@beck2001manifesto]). Our software framework makes this trade off by enforcing the use of consistent code naming conventions and file organisation which in turn enables automated generation of simple documentation at every code update. All model data-structures and algorithms are therefore always documented (at least minimally, with machine authored content), meaning model developers have a requirement to write customised documentation less frequently.

Reducing waste in research is a core responsibility of research funders [@chalmers2014increase] and funding the development of CHEMs that are not adequately understood, reused or updated is wasteful. Existing incentive structures for health economists generally do not promote facilitating peers to reuse their work and funding is rarely adequate to maintain an routinely update a model. Currently, it can take “an extraordinary amount of idealism” to commit to authoring and maintaining research software [@anzt2020environment]. Previously recommended strategies for more beneficial health economic research investments include support for harmonised ethical standards for model development [@10.3389/fpubh.2017.00068], methodological innovation to improve model transferability [@craig2018taking], networks of modellers working on common health conditions [@Sampson2019], and  centralized infrastructure such as open source model repositories [@Pouwels2022] and a standard platform for model implementations [@Ghabri2019]. Development of software frameworks to support ethical CHEM implementations could enable and enhance each of these strategies.

A future software framework for ethical CHEMs would ideally incorporate a base set of features useful to developers of computational models across all domains of public health, with the capability for community-led extensions that are tailored to the needs of modellers focused on specific health-conditions.

>>>>

<!-- Talk about DARTH (https://doi.org/10.1007/s40273-019-00837-x )/ BCEA -->
Ethical practice is a core expectation of health researchers and computational methods underpin most quantitative research, yet an understanding of what constitutes ethical public health modelling practice is underdeveloped [@10.3389/fpubh.2017.00068]. We have addressed this gap with our proposed project team responsibilities, enabling model attributes and model implementation assessment criteria.

@@

As illustrated by Table \@ref(tab:checktb), we have developed a prototype software framework that can be used to implement youth mental health CHEMs that largely satisfy our TRU criteria. However, in its current form our software framework is too fragile to be recommended as anything more than a prototype for supporting the development needs of other modelling teams and projects. We developed our prototype software framework primarily to address the user requirements of one group – ourselves. Factors such as user enjoyment, usability, active user-community and supporting resources influence adoption of software frameworks [@myllarniemi2018development]. We currently lack the resources required to provide ongoing address these issues.

However, our prototype has a number of features that future work on CHEM software frameworks may find useful to incorporate. Firstly, developing a software framework to work within an existing and widely used open source programming language such as R or python, can keep framework scope relatively narrow. This makes it more tractable to develop, maintain and learn, and can integrate other modelling tools written in that language (e.g., the dependency libraries we list in Table \@ref(tab:cpkgs)). Secondly, implementation that combines both object-oriented and functional programming paradigms can avail of the modular and syntactical simplicity benefits of the former, while limiting needless bundling of code artefacts. Thirdly, a sensible trade-off needs to be found between transparent code implementation (which requires clear and sufficiently detailed documentation) and Agile Software Development (for which a foundational principle is prioritising the development of working code over writing documentation [@beck2001manifesto]). Our software framework makes this trade off by enforcing the use of consistent naming conventions and file organisation which in turn enables automated generation of simple documentation at every code update. All model data-structures and algorithms are therefore always documented (at least minimally, with machine authored content), meaning model developers have a requirement to write customised documentation less frequently. Many of the features of our prototype software framework have the potential to be significantly enhanced through integration with evolving code-authoring and documentation authoring capabilities of large language models.

Our software framework provides the outline of a technical solution to some of the challenges of open source CHEM development. However, better tools alone are insufficient and policy and institutional changes are also required. Reducing waste in research is a core responsibility of research funders [@chalmers2014increase] and funding the development of CHEMs that are not adequately understood, reused or updated is wasteful. Existing incentive structures for health economists generally do not promote facilitating peers to reuse their work and funding is rarely adequate to maintain an routinely update a model. Currently, it can take “an extraordinary amount of idealism” to commit to authoring and maintaining research software [@anzt2020environment]. Previously recommended strategies for more beneficial health economic research investments include support for harmonised ethical standards for model development [@10.3389/fpubh.2017.00068], methodological innovation to improve model transferability [@craig2018taking], networks of modellers working on common health conditions [@Sampson2019], and  centralized infrastructure such as open source model repositories [@Pouwels2022] and a standard platform for model implementations [@Ghabri2019]. Development of software frameworks to support ethical CHEM implementations could enable and enhance each of these strategies.

A future software framework for ethical CHEMs would ideally incorporate a base set of features useful to developers of computational models across all domains of public health, with the capability for community-led extensions that are tailored to the needs of modellers focused on specific health-conditions. Such approaches will likely be important to ensure that the potential user-base is sufficient to make CHEM software frameworks sustainable.

### 2023-10-16 Markdown/Manuscript/Child_RMDs/MS_Imp_Example.Rmd
missing tables

<!-- Add intro text -->
We plan on using the ready4 software framework to implement multiple transparent, reusable and updatable CHEMs in youth mental health (see Figure \@ref(fig:fig1)).  Some initial outputs from this work are publicly available.

@@

## Case study: health and its value
We have previously describeda study [@Hamilton2021.07.07.21260129] to develop utility mapping models for use in samples of young people presenting to primary mental health services. The ready4 software framework was used in that study to develop CHEM modules, supply those modules with data and implement modelling analyses, creating the following artefacts:

>>>>

<!-- Add intro text -->
We are using the ready4 software framework to implement multiple transparent, reusable and updatable CHEMs in youth mental health (see Figure \@ref(fig:fig1)).  Some initial outputs from this work are publicly available as development version releases.

@@

## Case study: health and its value
We have previously described a study [@Hamilton2021.07.07.21260129] to develop utility mapping models for use in samples of young people presenting to primary mental health services. The ready4 software framework was used in that study to develop CHEM modules, supply those modules with data and implement modelling analyses, creating the following artefacts:

### 2023-10-16 Markdown/Manuscript/Child_RMDs/MS_Imp_Guidelines.Rmd
missing tables

We considered prior literature on modelling practice and our own professional experience to identify: (i) some core ethical responsibilities of CHEM developers and funders; (ii) attributes of CHEMs that can suggest fulfillment of these responsibilities; and (iii) criteria against which these attributes can be assessed.

## Ethical responsibilities of CHEM developers and funders
<!-- The ethical responsibilities and enabling model attributes we describe have both commonalities and distinctive features compared to a previous  -->
Boden and McKendrick [@10.3389/fpubh.2017.00068] proposed a framework for ethical computational modelling in public health based on the four ethical criteria of justice (concerning the social obligations of modellers to consider and communicate ethical issues), independence (concerning how modeller subjectivity shapes model design), transparency (concerning the need for policymakers to reliably evaluate model strengths and weaknesses) and beneficence / non-maleficence (concerning model quality and utility). However, the framework's authors mostly discuss these criteria in the context of a modelling project as a whole, including a model's conceptual and mathematical representations. The authors propose 13 questions to evaluate ethical risk across the four criteria, but of these only "is the model code open source or available on request" (for transparency) and "has the model been verified, i.e., does it do what the modeller wants it to do?" (for beneficence / non-maleficence) are specific to model's representation as software.

To reconceptualise these four criteria as ethical responsibilities that are specific to CHEM developers and funders, we first considered the distinct phases in a software development project. Software development lifecycle (SDLC) models take different approaches to the inclusion, naming, definition, sequencing and iteration of project phases [@ruparelia2010software]. However, SDLC models typically have components that can principally map to the concepts of *planning* (e.g. identification of system / user requirements and design specifications), *implementation* (e.g. development and testing) and *release* (e.g.deployment, system integration and maintenance). For simplicity, we propose one overarching ethical responsibility for CHEM developers and funders for each of these phases. These responsibilities are to:

(i) **make socially acceptable system requirements and design decisions** (during CHEM planning);

(ii) **ensure fitness for purpose** (of CHEM implementation); and

@@

Each responsibility combines different aspects of Boden and McKendrick's ethical criteria: responsibility (i) incorporates justice and independence; responsibility (ii) incorporates independence and transparency; and responsibility (iii) incorporates beneficence / non-maleficence and justice. Model developers and funders share these responsibilities as each one has implications for both professional practice and project resourcing.

@@

Misalignment between the values of computational model developers and those of the population groups affected by decisions based on their models presents significant ethical risks [@thompson2022escape; @duckett2022journey]. The value judgments of health economists influence the assumptions, selection of model features and standards for evidence that shape health economic model projects [@HARVARD2020112975]. These value judgments are rarely made explicit, omissions that may lead to socially unacceptable policy recommendations [@duckett2022journey]. modeller value judgments also determine the computational implementation of health economic models. For example, as illustrated in Table \@ref(tab:proscons), project resource constraints may require trade offs between desired CHEM **system requirements and design** features. These trade offs in turn determine how straightforward it is for CHEMs to be amended by third parties to reflect alternative value judgments in model conceptualisation and mathematical formalisation (Table \@ref(tab:timelygls)).

@@

Computational modellers have duties both to ensure computational model **fitness for purpose** and to provide potential third-party users with the means of assessing its adequacy for their proposed purposes [@Erdemir2020; @Feenstra2022; @thompson2019escape; @10.3389/fpubh.2017.00068]. However, it is common for health economic models to have serious methodological flaws [@carletto_zanuzzi_sammarco_russo_2020; @WONDER2015467]; insufficient validation [@Ghabri2019; @kolovos2017model; @haji2013model], poor reproducibility [@Jalali2021; @McManus2019; @Bermejo2017]; and undeclared errors [@Radeva2020]. Appropriate computational implementation choices can help address many of these shortcomings, for example by automating quality assurance checks and facilitating manual reviews by third parties.

@@

## CHEM attributes associated with ethical modelling practice
The responsibilities of model developers and funders over a CHEM lifecycle are easier to state than to measure.  However, aspects of ethical modelling practice may be inferred from measurable attributes of CHEMs. As described in Table \@ref(tab:timelygls),  we believe that the creators of transparent, reusable and updatable (TRU) CHEMs are likely to have fulfilled a number of their ethical obligations.  We therefore selected these model attributes to use as the basis for deriving ethical assessment criteria.

@@

Taking steps to make computational models **transparent** is an important part of ethical public health modelling practice [@10.3389/fpubh.2017.00068]. Guidance on transparency in health economic modelling recommended that model code and data should be clearly documented [@Eddy2012]. Notably, the same guidelines, published over ten years ago, did not recommend sharing model code and data. However, more recent healthcare modelling guidance does recommend public dissemination of such artefacts [@Erdemir2020]. Online repository services such as Zenodo [@Zenodo2013] and Dataverse [@Dataverse2007] provide persistent storage solutions that generate a Digital Object Identifier (DOI) for each code and data collection. Ensuring that calculations are correct and consistent with model specifications is an essential part of CHEM quality assurance [@techver2019]. The extensiveness of such verification checks can be reported using the concept of code coverage [@ERICWONG2010188] - the proportion of model code that has been explicitly tested. Tests should ideally combine both unit tests (to verify that small, isolated sections of code produce the correct output when run independently) and acceptance tests (to verify that the correct output is produced when multiple code components are run together to perform tasks that meet core user-requirements [@martin2003agile]). The nature and extent of individual model authorship contributions can become unclear when models are implemented over longer time-frames with a large and changing group of collaborators [@thompson2022escape]. This issue can be addressed by use of online repository services such as GitHub [@github2007], that provide citation tools and can transparently record all individual code contributions to a modelling project over its lifecycle.

@@

Models should be **updatable** so that they remain valid for longer, evolving as new evidence emerges and the systems being modelled change [@Jenkins2021; @garcia2021cost]. Ensuring that a model is regularly reviewed to identify and implement required improvements is a recommended defence against model validity decay [@calder2018computational]. Sustainable maintenance of open source research software requires both a core development team and an active user community [@info:doi/10.2196/20028]. Online communities can be an efficient means of engaging model users in testing each version of a model, identifying issues and suggesting improvements. Services such as GitHub [@github2007] provide collaborative code development tools [@MERGEL2015464] that help elicit, integrate and reconcile contributions from multiple contributors and to ensure each update is uniquely identifiable and retrievable. It is also important that verification checks are rerun with each model update, a task that can be automated using the software development practice of continuous integration [@CI2017].  The risk of model revisions having unintended consequences for third party users can be mitigated through the use of deprecation conventions [@zhou2016api] that take an informative and staged approach to retiring outdated model code and data.
<!-- Currently, the core development team for a CHEM will be typically be funded to produce a project end-point deliverable whose specifications are well defined early in the project. For more complex and multi-purpose CHEMs, particularly those designed to be incorporated into decision support systems, it may be better for development teams to adopt Agile Software Development [@beck2001manifesto], an approach that has been recommended for complex public health software projects [@10.3389/fpubh.2022.899874]. An Agile model will be less clearly specified in the initial project plan, but will instead continually develop in response to the requirements and feedback of users, who are provided with an initial, simplified working version of the model at the earliest feasible opportunity. -->

@@

-	**R2**: Terms of use place only ethical restrictions on reuse and allow anyone to ethically reuse model code and non-confidential data, in whole or in part, without charge, and for purposes that include the creation of derivative works.

>>>>

We considered prior literature on modelling practice and our own experience with developing and using CHEMs to identify: (i) some core ethical responsibilities of CHEM project teams; (ii) attributes of CHEMs that can suggest fulfillment of these responsibilities; and (iii) criteria against which these attributes can be assessed.

## Ethical responsibilities of CHEM project teams
<!-- The ethical responsibilities and enabling model attributes we describe have both commonalities and distinctive features compared to a previous  -->
Boden and McKendrick [@10.3389/fpubh.2017.00068] propose a framework for ethical public health modelling based on the criteria of independence (concerning how modeller subjectivity shapes model design), beneficence / non-maleficence (concerning model quality and utility), transparency (concerning the need for policymakers to reliably evaluate model strengths and weaknesses) and justice (concerning the social obligations of modellers to consider and communicate ethical issues about model use). However, these criteria are not specific to the computational model, but are designed to apply to a modelling project as a whole, including a model's conceptual and mathematical representations. The authors propose 13 questions to evaluate ethical risk across the four criteria, but of these, only "is the model code open source or available on request" (for transparency) and "has the model been verified, i.e., does it do what the modeller wants it to do?" (for beneficence / non-maleficence) are specific to model's representation as software.

To help identify ethical responsibilities that are specific to CHEM project teams, we first considered the distinct phases in a software development project. Software development lifecycle (SDLC) models take different approaches to the inclusion, naming, definition, sequencing and iteration of project phases [@ruparelia2010software]. However, SDLC models typically have components that can principally map to the concepts of *planning* (e.g., identification of system / user requirements and design specifications), *implementation* (e.g., development and testing) and *release* (e.g., deployment, system integration, maintenance and support). For simplicity, we propose one overarching ethical responsibility for CHEM developers and funders for each of these phases. These responsibilities are to:

(i) **establish socially acceptable system requirements and design specifications** (during CHEM planning);

(ii) **ensure fit for purpose implementation** (during CHEM implementation); and

@@

Each responsibility relates to aspects of Boden and McKendrick's ethical criteria. Transparency underpins all three responsibilities (Table \@ref(tab:timelygls). Additionally, the planning responsibility incorporates justice and independence, the implementation responsibility incorporates independence and the release responsibility incorporates beneficence / non-maleficence and justice. All three ethical responsibilities have implications for both professional practice and project resourcing and should therefore apply to model developers and funders jointly.

@@

Misalignment between the values of computational model developers and those of the population groups affected by decisions based on their models presents significant ethical risks [@thompson2022escape; @duckett2022journey].  The value judgments of health economic modellers are rarely adequately specified, omissions that may lead to socially unacceptable policy recommendations [@duckett2022journey]. These value judgments influence the assumptions, selection of model features and standards for evidence that shape health economic model projects [@HARVARD2020112975]. For example, as illustrated in Table \@ref(tab:proscons), project resource constraints may require trade-offs between desired CHEM **system requirements and design** features. These trade-offs in turn determine how straightforward it is for CHEMs to be amended by third parties to reflect alternative value judgments in model conceptualisation and mathematical formalisation (Table \@ref(tab:timelygls)).

@@

Modellers have duties both to ensure a computational model is **fit for purpose** and to provide potential third-party users with the means of assessing its adequacy for their proposed uses [@Erdemir2020; @Feenstra2022; @thompson2019escape; @10.3389/fpubh.2017.00068]. However, it is common for health economic models to have serious methodological flaws [@carletto_zanuzzi_sammarco_russo_2020; @WONDER2015467]; insufficient validation [@Ghabri2019; @kolovos2017model; @haji2013model], poor reproducibility [@Jalali2021; @McManus2019; @Bermejo2017]; and undeclared errors [@Radeva2020]. Appropriate computational implementation choices can help address many of these shortcomings, for example by automating quality assurance checks and facilitating manual reviews by third parties.

@@

## CHEM attributes associated with ethical modelling practice
The responsibilities of model project teams over a CHEM lifecycle are easier to state than to measure.  However, aspects of ethical modelling practice may be inferred from measurable attributes of CHEMs. As described in Table \@ref(tab:timelygls),  we believe that the creators of transparent, reusable and updatable (TRU) CHEMs are likely to have fulfilled a number of their ethical obligations.  We therefore selected these model attributes to use as the basis for deriving ethical assessment criteria.

@@

Taking steps to make computational models **transparent** is an important part of ethical public health modelling practice [@10.3389/fpubh.2017.00068]. Guidance on transparency in health economic modelling recommend that model code and data should be clearly documented [@Eddy2012]. Notably, the same guidelines, published over ten years ago, do not recommend sharing model code and data. However, more recent healthcare modelling guidance does recommend public dissemination of such artefacts [@Erdemir2020]. Online repository services such as Zenodo [@Zenodo2013] and Dataverse [@Dataverse2007] provide persistent storage solutions that generate a Digital Object Identifier (DOI) for each code and data collection. Ensuring that calculations are correct and consistent with model specifications is an essential part of CHEM quality assurance [@techver2019]. The extensiveness of such verification checks can be reported using the concept of code coverage [@ERICWONG2010188] - the proportion of model code that has been explicitly tested. Tests should ideally combine both unit tests (to verify that small, isolated sections of code produce the correct output when run independently) and acceptance tests (to verify that the correct output is produced when multiple code components are run together to perform tasks that meet core user-requirements [@martin2003agile]). The nature and extent of individual model authorship contributions can become unclear when models are implemented over longer time-frames with a large and changing group of collaborators [@thompson2022escape]. This issue can be addressed by use of online repository services such as GitHub [@github2007], that provide citation tools and can transparently record all individual code contributions to a modelling project over its lifecycle.

@@

Models should be **updatable** so that they remain valid for longer, evolving as new evidence emerges and the systems being modelled change [@Jenkins2021; @garcia2021cost]. Ensuring that a model is regularly reviewed to identify and implement required improvements is a recommended defence against model validity decay [@calder2018computational]. Sustainable maintenance of open source research software requires both a core development team and an active user community [@info:doi/10.2196/20028]. Online communities can be an efficient means of engaging model users in testing each version of a model, identifying issues and suggesting improvements. Services such as GitHub [@github2007] provide collaborative code development tools [@MERGEL2015464] that help elicit, integrate and reconcile contributions from multiple contributors and to ensure each update is uniquely identifiable and retrievable. It is also important that verification checks are rerun with each model update, a task that can be automated using the software development practice of continuous integration [@CI2017].  The risk of software revisions having unintended consequences for third party users can be mitigated through the use of deprecation conventions [@zhou2016api] that take an informative and staged approach to retiring outdated CHEMs.
<!-- Currently, the core development team for a CHEM will be typically be funded to produce a project end-point deliverable whose specifications are well defined early in the project. For more complex and multi-purpose CHEMs, particularly those designed to be incorporated into decision support systems, it may be better for development teams to adopt Agile Software Development [@beck2001manifesto], an approach that has been recommended for complex public health software projects [@10.3389/fpubh.2022.899874]. An Agile model will be less clearly specified in the initial project plan, but will instead continually develop in response to the requirements and feedback of users, who are provided with an initial, simplified working version of the model at the earliest feasible opportunity. -->

@@

-	**R2**: Terms of use place restrict only unethical usage and allow anyone to ethically reuse model code and non-confidential data, in whole or in part, without charge, and for purposes that include the creation of derivative works.

### 2023-10-16 Markdown/Manuscript/Child_RMDs/MS_Imp_Toolkit.Rmd
missing tables

Three framework libraries are designed to help streamline and standardise workflows for **authoring CHEM modules** from the template module. The R language supports functional and object-oriented programming paradigms [@10.1214/13-STS452]. Authoring with each of these paradigms is facilitated by a dedicated CHEM module authoring library. One module authoring library contains tools for writing functions in a consistent house style and then using the standardised naming conventions of that house style to automatically generate basic documentation for each function. A second module authoring library contains tools to help streamline and standardise the authoring and documenting of novel CHEM modules. A third module authoring library provides tools for disseminating themed bundles of CHEM modules as R libraries that are:

@@

We created a GitHub organisation (a collection of code repositories) where source code that we author is stored and version controlled. We configured individual repositories in our GitHub organisation to use GitHub actions to implement continuous integration. By default, code libraries authored with our framework will use continuous integration to assess compliance with policies specified by the Comprehensive R Archive Network (CRAN) [@CRAN2022]. To track our code coverage, we linked our GitHub organisation to an account we established at codecov [@codecov_2022]. To facilitate the creation and hosting of documentation websites, we enabled GitHub Pages in each repository used for code library development.

>>>>

Three framework libraries are designed to help streamline and standardise workflows for **authoring CHEM modules** from the template module. The R language supports functional and object-oriented programming paradigms [@10.1214/13-STS452]. Authoring with each of these paradigms is facilitated by a dedicated CHEM module authoring library. One module authoring library contains tools for writing functions in a consistent house style and automatically generating basic documentation for each function. A second module authoring library contains tools to help streamline and standardise the authoring and documenting of novel CHEM modules. A third module authoring library provides tools for disseminating themed bundles of CHEM modules as R libraries that are:

@@

We created a GitHub organisation (a collection of code repositories) where source code that we author is stored and version controlled. We configured individual repositories in our GitHub organisation to implement continuous integration. By default, code libraries authored with our framework will use continuous integration to assess compliance with policies specified by the Comprehensive R Archive Network (CRAN) [@CRAN2022]. To track our code coverage, we linked our GitHub organisation to an account we established at codecov [@codecov_2022]. To facilitate the creation and hosting of documentation websites, we enabled GitHub Pages in each repository used for code library development.

### 2023-10-16 Markdown/Manuscript/Child_RMDs/MS_Intro_Short.Rmd
missing tables

The set of digital files implementing a health economic model's structure, data and algorithms as software can be called a computational health economic model (CHEM). CHEMs can be developed using commercial modelling software or in an open-source programming language such as Python [@python2009] or R [@RCORE2022]. An advantage of commercial modelling software is that users often require no or limited programming skills to develop and apply CHEMs. However, open-source software programming languages may facilitate the development of CHEMs that are more transparent, reusable and updatable [@incerti2019r; @Pouwels2022]. The decision about what type of software development platform to use needs to balance ethical and pragmatic considerations  (see Table \@ref(tab:proscons)).

@@

Open-source CHEMs remain rare [@Feenstra2022; @Emerson2019]. Additional resources, skills and tools may be required to overcome a number of barriers to greater health economist adoption of open-source development. CHEM software development platforms were identified in a 2020 survey of health economists as a barrier to open-source model implementations  [@Pouwels2022]. Other reported barriers relate to difficulties in generalising models, updating models and transferring model data, concerns about the level of public access required and lack of interest from decision-makers, and legal, confidentiality and security considerations [@Pouwels2022].

<!-- A major consideration in open-source software development, is the selection of an appropriate software framework.  -->
<!-- on considerations such as model comprehensiveness, generalisability, reproducibility and transferability -->
<!-- Although there appears to be in principle support from many health economists for greater use of open source health economic models (OSHEMs)  [@Pouwels2022], actual implementations are rare [@Emerson2019; @Michalczyk2018; @Feenstra2022]. Barriers to adoption of OSHEMs include concerns about intellectual property, confidentiality, model misuse and the resources required to support open source implementations [@Pouwels2022; @Wu2019]. Adherence to good practice guidance is an essential requirement for healthcare modelling [@Erdemir2020],  -->

Better software frameworks could potentially improve open-source CHEM development. A software framework is a shared common technology used by developers to collaboratively author software and which is not typically visible to software end-users [@myllarniemi2018development]. A software framework provides a foundation for developing multiple software applications with shared resources (e.g., code and data files), that can be modified to suit specific needs. Advantages of using software frameworks include facilitating code reuse and extension, promoting good programming practice and the capability to provide enhanced functionality and performance without additional effort by developers [@edwin2014software]. Software frameworks have widely been developed and implemented in data science, for example, PyTorch [@PyTorch2019] for machine learning using Python.

@@

A high-level software framework for implementing open-source CHEMs in R has previously been developed with the primary aim of improving model transparency [@alarid2019need]. However, to gain widespread adoption, a software framework for CHEM developers may need to provide tools to help meet a broader range of ethical practice requirements. There are multiple challenges to overcome. Firstly, the ethical responsibilities of public health computational modellers remain poorly understood <!-- and inadequately fulfilled  [@pliakos2021ethics; @thompson2022escape;]-->[@10.3389/fpubh.2017.00068] and the computational modelling landscape is changing fast [@calder2018computational. Secondly, software frameworks can be challenging and time consuming to create and may become excessively complex over time [@edwin2014software]. Thirdly, software frameworks can be difficult to learn, often requiring model developers to undergo specialist training [@myllarniemi2018development].
<!-- a software framework is unlikely to be widely adopted unless the benefits to developers outweigh the costs associated with using it. -->
<!-- However, guidance on developing, selecting and using software frameworks for ethical CHEM implementations is scarce. -->

<!--  -->

<!-- Transparency has been proposed as one of the core ethical responsibilities of public health computational modellers [@10.3389/fpubh.2017.00068].  -->

<!-- Although computational models are widely used to inform health policy and system design [@dakin2015influence; @Erdemir2020; @LO2022100570; @Christene006827], -->

>>>>

The set of software files implementing a health economic model's structure, data and algorithms can be called a computational health economic model (CHEM). CHEMs can be developed using commercial modelling software or open-source programming languages such as Python [@python2009] or R [@RCORE2022]. An advantage of commercial modelling software is that users often require no or limited programming skills to develop and apply CHEMs. However, open-source software programming languages may facilitate the development of CHEMs that are more transparent, reusable and updatable [@incerti2019r; @Pouwels2022]. The decision about what type of software development platform to use must balance ethical considerations and resource constraints (see Table \@ref(tab:proscons)).

@@

Open-source CHEMs remain rare [@Feenstra2022; @Emerson2019]. The software used to develop CHEMs was identified as a barrier to open-source model implementations by participants in a 2020 survey of health economists [@Pouwels2022]. Other reported barriers relate to difficulties in generalising models, updating models and transferring model data, concerns about the level of public access and lack of interest from decision-makers, and legal, confidentiality and security considerations [@Pouwels2022]. 
<!-- Additional resources, skills and tools may be required to overcome a number of barriers to greater health economist adoption of open-source development. -->
<!-- A major consideration in open-source software development, is the selection of an appropriate software framework.  -->
<!-- on considerations such as model comprehensiveness, generalisability, reproducibility and transferability -->
<!-- Although there appears to be in principle support from many health economists for greater use of open source health economic models (OSHEMs)  [@Pouwels2022], actual implementations are rare [@Emerson2019; @Michalczyk2018; @Feenstra2022]. Barriers to adoption of OSHEMs include concerns about intellectual property, confidentiality, model misuse and the resources required to support open source implementations [@Pouwels2022; @Wu2019]. Adherence to good practice guidance is an essential requirement for healthcare modelling [@Erdemir2020],  -->

Better software frameworks could potentially help address some of the barriers to open-source CHEM development. A software framework is a shared common technology used by developers to collaboratively author software and which is not typically visible to software end-users [@myllarniemi2018development]. A software framework provides a foundation for developing multiple software applications with shared resources (e.g., code and data files), that can be modified to suit specific needs. Advantages of using software frameworks include facilitating code reuse and extension, promoting good programming practice and the capability to provide enhanced functionality and performance without additional effort by developers [@edwin2014software]. Software frameworks have widely been developed and implemented in data science, for example, PyTorch [@PyTorch2019] for machine learning using Python.

@@

A high-level software framework for implementing open-source CHEMs in R has previously been developed with the primary aim of improving model transparency [@alarid2019need]. However, to gain widespread adoption, a software framework for CHEMs may need to provide tools to help model developers meet a broader range of ethical practice requirements. There are multiple challenges to overcome. Firstly, the ethical responsibilities of public health computational modellers remain poorly understood <!-- and inadequately fulfilled  [@pliakos2021ethics; @thompson2022escape;]-->[@10.3389/fpubh.2017.00068] and the computational modelling landscape is changing fast [@calder2018computational]. Secondly, software frameworks can be challenging and time consuming to create and may become excessively complex over time [@edwin2014software]. Thirdly, software frameworks can be difficult to learn, often requiring model developers to undergo specialist training [@myllarniemi2018development].

<!-- a software framework is unlikely to be widely adopted unless the benefits to developers outweigh the costs associated with using it. -->
<!-- However, guidance on developing, selecting and using software frameworks for ethical CHEM implementations is scarce. -->
<!--  -->
<!-- Transparency has been proposed as one of the core ethical responsibilities of public health computational modellers [@10.3389/fpubh.2017.00068].  -->
<!-- Although computational models are widely used to inform health policy and system design [@dakin2015influence; @Erdemir2020; @LO2022100570; @Christene006827], -->
